WHOOMP! (There It Is) \\ Rapid Bayesian position reconstruction for gravitational\nobreakdashes-wave transients

### 2015-07-07 ms.tex
Add adaptive mesh example

Our sampling of the sky relies completely on the \acl{HEALPix} (\acs{HEALPix}), a special data structure designed for all\nobreakdashes-sky maps. \ac{HEALPix} divides the sky into equal\nobreakdashes-area pixels. There is a hierarchy of \ac{HEALPix} resolutions. A \ac{HEALPix} resolution may be designated by its order $N$. The $N=0$th order or base tiling has 12 pixels. At every successive order, each tile is subdivided into four new tiles. A resolution may also be referred to by the number of subdivisions along each side of the base tiles, $N_\mathrm{side} = 2^N$. There are $N^\mathrm{pix} = 12 {N_\mathrm{side}}^2$ pixels at any given resolution. The \ac{HEALPix} projection uniquely specifies the coordinates of the center of each pixel by providing a mapping from the resolution and pixel index $(N_\mathrm{side}, i_\mathrm{pix})$ to right ascension and declination $(\alpha, \delta)$.

We begin by evaluating the posterior probability density at the center of each of the $N_{\mathrm{pix},0} = 3072$ pixels of an $N_{\mathrm{side},0}=16$ \ac{HEALPix} grid. At this resolution, each pixel has an area of 13.4~deg$^2$. We then rank the pixels by contained probability (assuming constant probability density within a pixel), and subdivide the most probable $N_{\mathrm{pix},0}/4$ pixels into $N_{\mathrm{pix},0}$ new pixels. We then evaluate the posterior again at the pixels that we subdivided, sort again, and repeat seven times, so that we have evaluated the posterior probability density a total of $8 N_{\mathrm{pix},0}$ times. On \emph{most} iterations, we descend one level deeper in \ac{HEALPix} resolution. The resulting map is a tree structure that consists of a mix of several resolutions. This process is illustrated in Figure~\ref{fig:adaptive-illustration}.

We traverse the tree and flatten it into the highest resolution represented. The highest possible resolution is $N_\mathrm{side}=2^{11}$, with an area of $\approx 10^{-3}$~deg$^2$ per pixel.\footnote{Although the resulting sky map contains $N_\mathrm{pix} \approx 5\times10^6$ pixels, at most $\approx 2\times10^4$ pixels have distinct values. For the purpose of delivery to observers, therefore, the output is always \texttt{gzip}\nobreakdashes-compressed with a ratio of $\approx 250:1$.}

@@

@@

Since \ac{BAYESTAR} is designed as one of the final steps in the real\nobreakdashes-time \ac{BNS} search, it is important to characterize how long it takes to calculate a sky map. We compiled \ac{BAYESTAR} with the Intel~C~Compiler~(\texttt{icc}) at the highest architecture\nobreakdashes-specific optimization setting (\texttt{-xhost}). We timed it under Scientific~Linux~6 on a Supermicro~1027GR\nobreakdashes-TRFT system with dual Intel Xeon E5-2670 CPUs clocked at 2.60GHz, providing a total of 16 cores and capable of executing 32 threads simultaneously (with hyperthreading). In Figure~\ref{fig:runtimes}, we show how long it took to calculate a localization with \ac{BAYESTAR} as the number of OpenMP threads was varied from 1 to 32. This is a violin plot, a smoothed vertical histogram. The red surface shows run times for a two\nobreakdashes-detector network (HL) modeled on the first scheduled Advanced \ac{LIGO} observing run in 2015, and the blue surface shows run times for a three\nobreakdashes-detector network (HLV) based on the second planned observing run in 2016. These are the two observing scenarios that are discussed in \cite{FirstTwoYears}.

>>>>

Our sampling of the sky relies completely on the \acl{HEALPix} (\acs{HEALPix}), a special data structure designed for all\nobreakdashes-sky maps. \ac{HEALPix} divides the sky into equal\nobreakdashes-area pixels. There is a hierarchy of \ac{HEALPix} resolutions. A \ac{HEALPix} resolution may be designated by its order $N$. The $N=0$th order or base tiling has 12 pixels. At every successive order, each tile is subdivided into four new tiles. A resolution may also be referred to by the number of subdivisions along each side of the base tiles, $N_\mathrm{side} = 2^N$. There are $N_\mathrm{pix} = 12 {N_\mathrm{side}}^2$ pixels at any given resolution. The \ac{HEALPix} projection uniquely specifies the coordinates of the center of each pixel by providing a mapping from the resolution and pixel index $(N_\mathrm{side}, i_\mathrm{pix})$ to right ascension and declination $(\alpha, \delta)$.

The \ac{BAYESTAR} adaptive sampling process works as follows. W begin by evaluating the posterior probability density at the center of each of the $N_{\mathrm{pix},0} = 3072$ pixels of an $N_{\mathrm{side},0}=16$ \ac{HEALPix} grid. At this resolution, each pixel has an area of 13.4~deg$^2$. We then rank the pixels by contained probability (assuming constant probability density within a pixel), and subdivide the most probable $N_{\mathrm{pix},0}/4$ pixels into $N_{\mathrm{pix},0}$ new daughter pixels. We then evaluate the posterior again at the centers of the new daughter pixels, sort again, and repeat seven times. After the last iteration, we have evaluated the posterior probability density a total of $8 N_{\mathrm{pix},0}$ times. On \emph{most} subdivision steps, we descend one level deeper in \ac{HEALPix} resolution. This process is illustrated in Figure~\ref{fig:adaptive-illustration}.

@@

The resulting map is a tree structure that describes a mesh of pixels with different resolutions. An example \ac{BAYESTAR} subdivision is shown in Figure~\ref{fig:adaptive-illustration}. To convert this mesh into a FITS image, 
we traverse the tree and flatten it into the highest resolution represented. The highest possible resolution is $N_\mathrm{side}=2^{11}$, with an area of $\approx 10^{-3}$~deg$^2$ per pixel.\footnote{Although the resulting sky map contains $N_\mathrm{pix} \approx 5\times10^6$ pixels, at most $\approx 2\times10^4$ pixels have distinct values. For the purpose of delivery to observers, therefore, the output is always \texttt{gzip}\nobreakdashes-compressed with a ratio of $\approx 250:1$.}

\begin{figure}
    \includegraphics[width=\columnwidth]{adaptive_mesh}
    \caption{\label{fig:adaptive-mesh}}
\end{figure}

@@

Since \ac{BAYESTAR} is designed as one of the final steps in the real\nobreakdashes-time \ac{BNS} search, it is important to characterize how long it takes to calculate a sky map. We compiled \ac{BAYESTAR} with the Intel~C~Compiler~(\texttt{icc}) at the highest architecture\nobreakdashes-specific optimization setting (\texttt{-xhost}). We timed it under Scientific~Linux~6 on a Supermicro~1027GR\nobreakdashes-TRFT system with dual Intel Xeon E5-2670 CPUs clocked at 2.60GHz, providing a total of 16 cores and capable of executing 32 threads simultaneously (with hyperthreading). In Figure~\ref{fig:runtimes}, we show how long it took to calculate a localization with \ac{BAYESTAR} as the number of OpenMP threads was varied from 1 to 32. This is a violin plot, a smoothed vertical histogram. The magenta regions show run times for a two\nobreakdashes-detector network (HL) modeled on the first scheduled Advanced \ac{LIGO} observing run in 2015, and the blue regions show run times for a three\nobreakdashes-detector network (HLV) based on the second planned observing run in 2016. These are the two observing scenarios that are discussed in \cite{FirstTwoYears}.
